### 2019-10-11 main_text/main_text.tex
A few more edits to intro

Through physical and mathematical modeling, we investigate a more complex
but common switch system where the bistability is due, as we will show, to a
membrane transport protein. Its key feature is the indirect activation of
the transporter gene by the transporter substrate, leading to positive
feedback similar to the aforementioned ``on-off'' switch. An example for
such an architecture is the \emph{lac} operon, where lactose indirectly
activates the expression of lactose permease. Other examples in \emph{E.
coli} include the \emph{araFGH} and \emph{xylFGH} genes, which encode for
arabinose and xylose transporters, respectively. For \emph{lac} and
\emph{araFGH}, bistability has indeed been observed and attributed to such a
positive feedback
loop~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
A eukaryotic example is the glucose transporter GLUT-2 in liver and
\textbeta-cells~\cite{Bae2010,Tiedge1991}, though this system is much more
complex than the following analysis.

It is quite conceivable that this auto-activation process is common to many
substances that a cell would want to accumulate. Such a switch enables the
cell to sense and respond to its environment: if the substrate enters the
cell, it activates the production of membrane transporters. The cell then
starts accumulating the substrate, thereby ``testing'' the substrate's
presence in the extracellular environment. If there is enough, the
expression stabilizes at an ``on'' point and the cell has, in a short-term
sense, adapted. When there is not enough substrate anymore, the gene, which
often encodes for a whole set of enzymes for this one metabolite, switches
``off'' again. Such a mechanism could be involved in various cases of
short-term adaptation.

\textit{MM: I like the above paragraph, but we should also add something
about the TF \& frame it as resourse utility: xapR, lacI, and I suspect
others of these TFs happen to be constitutively expressed at low copy
number. They act as ``always on'' sensors that trigger a switch to produce
massive numbers of metabolic enzymes only when their substrate is
available.}

For our modeling, we focus on the \emph{xapABR} genetic circuit from
\emph{E. coli} as a case study. It is similar to \emph{lac}, but less
complex. Instead of lactose, its purpose is to make use of the nucleoside
xanthosine as an energy source~\cite{Buxton1980,Hammer-Jespersen1980}. The
circuit is made up of two operons: one that encodes for XapR and another
that encodes for XapA and XapB. XapR appears to be a transcription factor
that is induced by xanthosine and activates the \emph{xapAB} promoter.
This promoter seems to have two binding sites for XapR, but the promoter
architecture and function is not yet fully understood. The
transcription of \emph{xapR} seems to be constitutive and not
auto-regulated~\cite{Seeger1995}. Structural homology to other transcription
factors suggests that XapR appears in dimers where one dimer can bind two
xanthosine molecules~\cite{Joergensen1999}. The protein XapA is a purine
nucleoside phosphorylase that degrades xanthosine into components (ribose
and xanthine) that can be fed into metabolic
pathways~\cite{Buxton1980,Hammer-Jespersen1980}. XapB on the other hand is a
membrane transporter of xanthosine~\cite{Seeger1995,Norholm2001}.

>>>>

Through physical and mathematical modeling, we investigate a more complex
but common switch system where the bistability is due, as we will show, to a
membrane transport protein. Its key feature is the indirect activation of
the transporter gene by the transporter substrate, leading to positive
feedback similar to the aforementioned ``on-off'' switch. An example for
such an architecture is the \emph{lac} operon, where lactose indirectly
activates the expression of lactose permease. Other examples in \emph{E.
coli} include the \emph{araFGH} and \emph{xylFGH} operons, which contain genes for
arabinose and xylose transporters, respectively. For \emph{lac} and
\emph{araFGH}, bistability has indeed been observed and attributed to such a
positive feedback
loop~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
A eukaryotic example is the glucose transporter GLUT-2 in liver and
\textbeta-cells~\cite{Bae2010,Tiedge1991}, though this system is much more
complex than the following analysis.

It is quite conceivable that this auto-activation process is common to many
substances that a cell would want to accumulate. Such a switch enables the
cell to sense and respond to its environment: if the substrate enters the
cell, it activates the production of membrane transporters. The cell then
starts accumulating the substrate, thereby ``testing'' the substrate's
presence in the extracellular environment. If there is enough, the
expression stabilizes at an ``on'' point and the cell has, in a short-term
sense, adapted. When there is not enough substrate anymore, the operon, which
often encodes for a whole set of enzymes for this one metabolite, switches
``off'' again. Such a mechanism could be involved in various cases of
short-term adaptation.

A key element of this mechanism is the presence of a transcription factor
which is always expressed at a low level (often at copy numbers
of order $\sim10$) and which binds to the transporter substrate.
This is resource efficient for the cell, as this low copy number
transcription factor acts as an ``always on'' sensor to detect
the substrate, allowing high copy numbers of the membrane transporter
and its attendant operon to be expressed only when
their substrate is actually present. The transcription factors
LacI, AraC, and XylR all appear to fill this
role~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
\todo{do we have a cite for xylR too?}

For our modeling, we focus on the \emph{xapABR} genetic circuit from
\emph{E. coli} as a case study. It is similar to \emph{lac}, but less
complex. Instead of lactose, its purpose is to make use of the nucleoside
xanthosine as an energy source~\cite{Buxton1980,Hammer-Jespersen1980}. The
circuit is made up of two operons: one that encodes for XapR and another
that encodes for XapA and XapB. XapR appears to be a transcription factor
that is induced by xanthosine and activates the \emph{xapAB} promoter,
in close analogy to LacI, AraC, and XylR.
The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
but the promoter architecture and function is not yet fully understood. The
transcription of \emph{xapR} seems to be constitutive and not
auto-regulated~\cite{Seeger1995}. Structural homology to other transcription
factors suggests that XapR appears in dimers where one dimer can bind two
xanthosine molecules~\cite{Joergensen1999}. The protein XapA is a purine
nucleoside phosphorylase that degrades xanthosine into components (ribose
and xanthine) that can be fed into metabolic
pathways~\cite{Buxton1980,Hammer-Jespersen1980}. XapB on the other hand is a
membrane transporter of xanthosine~\cite{Seeger1995,Norholm2001}.
